---
title: "Quantizing Mariculture Metrics"
author: "Owen Liu"
date: "September 26, 2016"
output: pdf_document
---

THIS IS A CLONE OF THE NORMALIZATION SCRIPT, BUT WITH QUANTILES INSTEAD OF ZERO TO 1 NORMALIZATION
    
```{r global options,echo=F}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

### Selecting and Joining Relevant Variables

Import the data

```{r setup, include=FALSE,message=F}
library(dplyr)
library(Hmisc)
W_D <- getwd()
dat<- read.csv(file=paste0(W_D,"/data/data_full_prenorm_101716.csv"))
```

Select relevant variables
```{r choose vars}
dat2 <- dat %>%
  select(country,year,
         Q.balance,UV.balance,energy_adequacy,
         gdp,calories:polyunsatFA,
         polyunsatFA_percentseafood:prod_ratio,
         aq_gdp)
```

### Variable Quantiles

All variables will be broken into quintiles (5 categories).

#### Trade metrics


```{r trade}
quantize <- function(x) {
  cats<-Hmisc::cut2(x,g=5)
  as.numeric(cats)
}
# compare distribution of raw/quantized data, for Q.balance
test <- quantize(dat2$Q.balance)
par(mfrow=c(2,1))
# hist(dat2$Q.balance, xlab="Raw Trade Balance (Exports/Imports in tonnes)",main="",breaks=20)
# hist(as.numeric(test), xlab="Normalized Trade Balance Score")

# Looks good
dat.quant <- mutate(dat2,Q.balance_q=quantize(Q.balance),UV.balance_q=quantize(UV.balance))
```

***

Production balance, prod_ratio, of aquaculture to fisheries production

```{r norm production}
# norm_prod_ratio <- function(x) {
#   nrm <- log10(x)
#   out <- (nrm-min(nrm[is.finite(nrm)]))/(max(nrm[is.finite(nrm)])-min(nrm[is.finite(nrm)]))
#   out[is.infinite(nrm)] <- 0
#   return(out)
# }

dat.quant <- mutate(dat.quant,prod_ratio_q=quantize(prod_ratio))
```

Not sure how much I like this one (normalization on top of log transformation really distorts the scale to where a country with still a very small production ratio).

***

Total number of species harvested, relative to the most observed.

```{r num species}
#normalize <- function(x) (x-min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T))
dat.quant <- mutate(dat.quant,spp_farmed_q=quantize(spp_farmed))

```

***

#### Food Security Data

energy_adequacy, scaled relative to maximum in the data

```{r energy ad}
# DEFUNCT FOR NOW 10/11/16
# norm_energy_ad <- function(x) {
#   x[x>100]<-100
#   out <- normalize(x)
#   return(out)
# }
dat.quant <- mutate(dat.quant,energy_adequacy_q=quantize(energy_adequacy))
```


***

GDP per capita, scaled relative to richest country.

```{r gdp}
dat.quant <- mutate(dat.quant,gdp_q=quantize(gdp))
# hist(dat.quant$gdp,xlab="quantized GDP Score",main="")
```

Aquaculture as percent of GDP

```{r aq gdp}
dat.quant <- mutate(dat.quant,aq_gdp_q=quantize(aq_gdp))
```

***

#### Nutrition Data

We first need to combine all of the vitamins into one metric. We do this by normalizing them individually and then taking a mean for each country across the individual scores. Also quantize calories, proteins, FAs

```{r vitamins}
# Scale each vitamin
dat.quant <- dat.quant %>% mutate_each(funs(q=quantize(.)),calories:zinc_percentseafood)

# Calculate a mean vitamin score
dat.quant$vitamins_all <- dat.quant %>% 
  select(vitaminA_percentseafood_q:zinc_percentseafood_q) %>% 
  rowMeans(na.rm=TRUE)

dat.quant$vitamins_all[is.nan(dat.quant$vitamins_all)] <- NA
```

***


#### Ecological Data

All of the ecological metrics are scaled to 0 to 10. We just redefine this to a 0 to 1 scale.

```{r ecological data}
dat.quant <- dat.quant %>% mutate_each(funs(q=quantize(.)),native:habitat)
```

***

#### Averages

First-cut means across the categories

```{r }
# gm_mean = function(x, na.rm=FALSE){
#   exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
# }

dat.quant$trade <- dat.quant %>% select(Q.balance_q,UV.balance_q,prod_ratio_q,aq_gdp_q) %>% apply(.,1,FUN=mean)
dat.quant$fs <- dat.quant %>% select(energy_adequacy_q,gdp_q) %>% apply(.,1,FUN=mean)
dat.quant$reliance <- dat.quant %>% select(polyunsatFA_percentseafood_q:zinc_percentseafood_q) %>% apply(.,1,FUN=mean)
dat.quant$nutrition <- dat.quant %>% select(calories_q:polyunsatFA_q) %>% apply(.,1,FUN=mean)
dat.quant$ecol <- dat.quant %>% select(native_q:habitat_q) %>% apply(.,1,FUN=mean)

## Overall mean (excluding ecological score)
dat.quant$totscore <- dat.quant %>% select(trade:nutrition) %>% apply(.,1,FUN=mean)
```

### Output data

Write output

```{r output}
write.csv(dat.quant,file=paste0(W_D,"/data/data_quantized_101716.csv"))
```

Graph of country scores along different axes

```{r graph}
summary_dat <- dat.quant %>% select(country,trade:nutrition)

ggplot(summary_dat,aes(x=trade,y=nutrition)) +
  geom_point()+
  geom_text(aes(label=country),size=2,vjust=1.5)+
  xlim(5,1)+ylim(5,1)+
  theme_minimal()+
  xlab("Economic Potential")+
  ylab("Nutrition Potential")
ggsave("econ_nutrtion_scores.png")
ggplot(summary_dat,aes(x=reliance,y=nutrition)) +
  geom_point()+
  geom_text(aes(label=country),size=2,vjust=1.5)+
  xlim(1,5)+ylim(5,1)+
  theme_minimal()+
  xlab("Reliance on Seafood")+
  ylab("Nutrition Potential")
ggsave("reliance_nutrition_scores.png")
```