---
title: "master_dataset_construction"
author: "Owen Liu"
date: "May 12, 2016"
output: html_document
---
## Combining Nutritional, Ecological, and Economic Datasets to produce a "Master" global mariculture dataset

The idea is to produce two different datasets, one organized by species with their associated nutritional, ecological, and economic data, and one organized by countries, with their associated nutritional/poverty status, ecological scores, and economic parameters


```{r dataset importing}
library(plyr)
library(dplyr)
library(RColorBrewer)
library(ggplot2)
lunique <- function(x) length(unique(x))

# Working directory
W_D <- getwd() # Owen's working directory where all the scripts are.  Data files are in this wd/data

# Renato's value, aquaculture production, and world trade datasets
value <- read.csv(file=paste0(W_D,"/data/WORLD_VALUE.csv"))
prod <- read.csv(file=paste0(W_D,"/data/WORLD_AQ_PRODUCTION.csv"))
trade <- read.csv(file=paste0(W_D,"/data/WORLD_TRADE.csv"))

# Trujillo ecological data
# By country
truj.coun <- read.csv(file=paste0(W_D,"/data/trujillo_countries.csv"))
# Fix the names (for descriptions of variables, see Trujillo 2008 in Alder and Pauly, 2008)
names(truj.coun) <- c("country","native","ex.imp","fishmeal","intensity","nutri","hatchery","antibiotic","habitat","GMO","code","trace","employment","waste")

# By species
truj.spe <- read.csv(file=paste0(W_D,"/data/trujillo_species.csv"))
# Fix the names
names(truj.spe) <- c("country","species","native","fishmeal","intensity","hatchery","habitat","waste","ecol.score","msi")

# Nutritional status by country
foodsecurity <- read.csv(file=paste0(W_D,"/data/Food Security Data/FOODSECURITY.csv"))
```

Now we can join the various data sets into two Master datasets; one organized by country/year/species for species-specific variables, and one by country/year for country-specific (non species-specific) variables

```{r join data}
# Join value and production data and remove ID columns
econ <- full_join(value,prod,by=c("species","year","country","area"))
econ <- econ[,-c(1,9)]

# Join Trujillo species data to econ dataset
spe.data <- full_join(econ,truj.spe,by=c("country","species"))

# Join nutritional status data to trade and Trujillo country data and remove excess columns
coun.data <- full_join(trade,foodsecurity,by=c("country","year"))
coun.data <- coun.data[,-c(1,11)]
coun.data <- full_join(coun.data,truj.coun,by="country")

# Save data
write.csv(spe.data,file=paste0(W_D,"/data/species_data_full.csv"))
write.csv(coun.data,file=paste0(W_D,"/data/country_data_full.csv"))

```

Rows (countries or species) with "complete" data (data from all sources for a given dataset), just given the data we have:

```{r complete cases}
coun.data.comp <- coun.data[complete.cases(coun.data),]
spe.data.comp <- spe.data[complete.cases(spe.data),]
write.csv(spe.data.comp,file=paste0(W_D,"/data/species_data_complete_cases.csv"))
write.csv(coun.data.comp,file=paste0(W_D,"/data/country_data_complete_cases.csv"))

```

Rows with complete data for high priority variables (5.25.16)

```{r high priority}
high.vars <- c("year","country","species","area","value","value.kg","relative.value","production","energy_adequacy","undernourish","total_population","protein","gdp","Q.exp","Q.imp","V.exp","V.imp","Q.balance","V.balance","UV.balance","native","fishmeal","hatchery","habitat","waste","ecol.score")

coun.data.high <- select(coun.data,which(names(coun.data) %in% high.vars)) %>% na.omit()
spe.data.high <- select(spe.data,which(names(spe.data) %in% high.vars)) %>% na.omit()

```

We can work with these data for now.  Some rankings and summary statistics:

```{r preliminary data viz}
# Number of species by country in the dataset
num.spe <- spe.data.high %>% group_by(country) %>% summarise(numspe=n_distinct(species)) %>% arrange(desc(numspe))

num.spe.all <- spe.data %>% group_by(country) %>% summarise(numspe=n_distinct(species)) %>% arrange(desc(numspe))


# Countries by max production exported (max exports by country, regardless of year)
# Countries by max value exported (max exports by country, regardless of year)
# Countries by largest AVERAGE trade surplus in production
# Countries by largest AVERAGE trade surplus in value
# Countries by largest AVERAGE surplus in value PER UNIT

coun.stats <- coun.data.high %>% group_by(country) %>% summarise(Qexp.max=max(Q.exp),Vexp.max=max(V.exp),Qbal.mean=mean(Q.balance),Vbal.mean=mean(V.balance),UVbal.mean=mean(UV.balance))

```

Plotting country ranks

```{r country plots}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

Qexp.max.plot <- ggplot(arrange(coun.stats,desc(Qexp.max)),aes(x=country,y=Qexp.max))+geom_bar(stat="identity") + coord_flip()

```